---
title: "ENVS-193DS_seed-counts"
author: Lena Hicks, Jasmine Toni, Izzy Tector
date: June 15 2023
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: yeti
editor: visual
execute: 
  message: false
  warning: false
---

# Problem 2. Seed Counts

## Introduction

In the realm of biological research, understanding the factors that influence seed count is of great significance as it sheds light on the reproductive success and fitness of plant species. In the alpine tundra, results of climate change have included 'shrubification'- an increase in woody shrubs, which affects on the vegetative community are in need of inspection([Mod & Luoto, 2016](https://iopscience.iop.org/article/10.1088/1748-9326/11/12/124028/meta)). This analysis aims to investigate how seed count varies in relation to three key variables: plot type (shrub or open), plant species, and the total number of inflorescences, and to see if a simpler model explaining seed count can be derived. Previous studies have shown shrubs to increase seed bank count in desert areas and that different plant species display varying seed production patterns, where trade-offs may occur between larger seed mass and seed count, which are interesting points to compare to the alpine tundra environment we studied([Filazzola et. al., 2019](https://doi.org/10.1371/journal.pone.0215988); [Paul-Victor & Turnbull, 2009](https://doi.org/10.1371/journal.pone.0006917)). Our null hypothesis is that none of the variables observed, plot type, species, nor number of inflorescences, will have an effect on seed count, and our alternative hypothesis is that at least one of these variables will have a correlation with seed count.

## Methods

```{r loading-packages}
library(MASS) # have to read this in before tidyverse 
# should haves
library(tidyverse)
library(here)
library(janitor)
library(ggeffects)
library(performance)
library(naniar)
library(skimr)
library(flextable)
library(car)
library(broom)
library(corrplot)
library(AICcmodavg)
library(GGally)
# would be nice to have
library(MuMIn)
library(DHARMa)
#library(equatiomatic)
library(corrplot)
library(lme4)
library(glmmTMB)
```

```{r loading-data}
seed_count <- read_csv(here("data", "shrubstudy_seed_ctwt.ms.data.csv")) %>% 
  # make the column names cleaner
  clean_names() %>% 
  select(treatment, species, total_nr_infl, nr_infl_coll, nr_seeds)
```

```{r data-vis}
ggplot(seed_count, aes(x = nr_seeds)) +
  geom_histogram(bins = 17)+
  labs(x = "Seed Count", y = "Number of Data Samples Taken", title = "Amount of Seeds Counted per Each Data Sample Taken", hjust = 0.5)
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 1. Seed Count per Data Sample visualization.** Histogram displaying how many seeds were counted at each data sample. The x axis is the seed count and the y axis is amounted of data samples collected per each seed count. Right-skewed distribution is observed, indicating many samples had little to no seeds, with some outliers of data samples having very high seed counts.
:::

Checking Missing Data:

```{r vis-missing-data}
gg_miss_var(seed_count)
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 2. Visualizing Missing Data.** Checking visually the missing data and where it occurs. Seed count has the most missing data, with 106 samples missing. Inflorescence collected has 6 samples missing, and the rest of the data is not missing.
:::

```{r removing-missing-data}
# creating subset of data with only total collected data included
seed_subset <- seed_count %>% 
  # dropping NA's from data set
  drop_na(treatment, species, total_nr_infl, nr_infl_coll, nr_seeds)
```

```{r pairs-plot}
seed_subset %>% 
  select(!species) %>% 
  ggpairs()
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 3. Pairs Plot of Variables Relationships.** Pairs plot used to display the individual relationships between variables and their correlation to each other. Potentially positive linear relationship appears between all of the variables, each with high correlation above 0.9.
:::

```{r skim-data}
skim(seed_subset)
```

### Building Models

```{r building-models}
# linear model, we know this is wrong
seedmod1 <- lm(nr_seeds ~ treatment + species + total_nr_infl, data = seed_subset)

# generalized linear model with Poisson distribution
seedmod2 <- glm(nr_seeds ~ treatment + species + total_nr_infl, data = seed_subset, family = "poisson")
seedmod2.a <- glm(nr_seeds ~ treatment + species + total_nr_infl, data = seed_subset, family = "poisson")

# generalized linear model with negative binomial distribution
seedmod3 <- glm.nb(nr_seeds ~ treatment + species + total_nr_infl, data = seed_subset)
seedmod3.a <- glmmTMB(nr_seeds ~ treatment + species + total_nr_infl, data = seed_subset, family = "nbinom2")

# generalized linear model with Poisson distribution and random effect of site
seedmod4 <- glmer(nr_seeds ~ species + total_nr_infl + (1|treatment), data = seed_subset, family = "poisson")
seedmod4.a <- glmmTMB(nr_seeds ~ species + total_nr_infl + (1|treatment), data = seed_subset, family = "poisson")

# generalized linear model with negative binomial distribution and random effect of site
seedmod5 <- glmer.nb(nr_seeds ~ species + total_nr_infl + (1|treatment), data = seed_subset)
seedmod5.a <- glmmTMB(nr_seeds ~ species + total_nr_infl + (1|treatment), data = seed_subset, family = "nbinom2")
```

```{r check-assumptions-mod-1}
# check diagnostics
plot(simulateResiduals(seedmod1)) # bad
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 4. Assumptions Test of Model 1.** Dharma test of residuals for model 1. Model appears bad due to extreme deviation.
:::

```{r check-assumptions-mod-2}
plot(simulateResiduals(seedmod2)) # bad
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 5. Assumptions Test of Model 2.** Dharma test of residuals for model 2. Model appears bad due to extreme deviation.
:::

```{r check-assumptions-mod-3}
plot(simulateResiduals(seedmod3)) # ok?
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 6. Assumptions Test of Model 3.** Dharma test of residuals for model 3. Model has fairly straight line in Q-Q plot suggesting potential normality.
:::

```{r check-assumptions-mod-4}
plot(simulateResiduals(seedmod4)) # bad
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 7. Assumptions Test of Model 4.** Dharma test of residuals for model 4. Model appears bad due to extreme deviation.
:::

```{r check-assumptions-mod-5}
plot(simulateResiduals(seedmod5)) # bad/ok?
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 8. Assumptions Test of Model 5.** Dharma test of residuals for model 5. Model appears okay, the data is nearing normality in Q-Q plot, but not best model prediction.
:::

```{r AICc-assumptions}
MuMIn::model.sel(seedmod1, seedmod2, seedmod3, seedmod4, seedmod5)
```

```{r}
seedmod3
```

```{r}
summary(seedmod3)
```

```{r}
confint(seedmod3)
```

```{r}
# adjusted R2
r.squaredGLMM(seedmod3)
```

```{r null-model-comparison}
# Overall test of the model's significance
null_model <- glm.nb(nr_seeds ~ 1, data = seed_subset) 
# Null model with only intercept
anova(seedmod3, null_model, test = "Chisq")
```

```{r simple-model-vis-treatment}
plot(ggpredict(seedmod3, terms = "treatment", back.transform = TRUE), add.data = TRUE) +
  labs(title = "Seed Count Using Only Treatment as Predictor:\n Shrub vs Open(control)", x = "Treatment Type", y = "Seed Count", hjust = 0.5)
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 9. Treatment as Simple Model for Seed Count.** Visually assessing using a simple model of only treatment type predictor of seed count. Overall, both shrub and control(open) are more clustered towards lower seed counts, but open appears to be a little more evenly spread in having higher seed counts.
:::

```{r simple-model-vis-species}
plot(ggpredict(seedmod3, terms = "species", back.transform = TRUE), add.data = TRUE)+
  labs(title = "Seed Count Using Only Species as Predictor", x = "Species", y = "Seed Count")
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 10. Species as Simple Model for Seed Count.** Visually assessing using simple model of just species as predictor to seed count. Species appears to have large effect on seed count with high variabilty shown amongst different species.
:::

```{r simple-model-vis-inflorescence}
plot(ggpredict(seedmod3, terms = "total_nr_infl", back.transform = TRUE), add.data = TRUE)+
  labs(title = "Inflorescences Counted as Simple Predictor\n to Seed Count", x = "Inflorescences Counted", y = "Seed Count")
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 11. Inflorescence as Simple Model for Seed Count.** Visually assessing using simple model of just inflorescence counted as predictor to seed count. Exponential relationship is displayed, with increased inflorescence count having a drastic effect on seed count.
:::

```{r simpler-models}
# create models with only one variable as the predictor to test if a simpler model will accurately predict seed count
seedmod_treatment <- lm(nr_seeds ~ treatment, data = seed_subset)
seedmod_species <- lm(nr_seeds ~ species, data = seed_subset)
seedmod_total_nr_infl <- lm(nr_seeds ~ total_nr_infl, data = seed_subset)
```

```{r AICc-simple-models}
MuMIn::model.sel(seedmod3,null_model,seedmod_treatment,seedmod_species,seedmod_total_nr_infl)
```

## Results

The results of doing a Chi Square test on the seedmod3 model (created with generalized linear model with negative binomial distribution) it was found that the seedmod3 was the best model of the five that were tested as the Chi-square value using a significance value of 0.05 for the seedmod3 was 178.6161 as seen in the ANOVA table. The degrees of freedom were 181 for the null model and 174 for the full model, the p-value for the full model was indicated to be \<0.001 in the ANOVA table.

Then, based on seedmod3, simpler models were tested with predictors of only one of the variables instead of the full model (seedmod3) like seedmod_treatment. Comparing these models showed that the best fit model to the data based on the lowest AICc value was the seedmod3. The AICc value for the simple models (total_nr_infl = 1295.6, species = 1722.1, treatment = 1741.6) were all larger than the seedmod3's AICc value (1132.0). This showed that there was no simpler model for this data frame to predict the seed count based on the AICc values.

### Visualization

```{r results-vis}
# create a ggplot with species and treatment as predictors of seed count
ggplot(data = seedmod3, aes(x = species, y = nr_seeds, fill = treatment)) +
  geom_boxplot(position = position_dodge(width = 0.8), color = "black") +
  labs(x = "Species", y = "Seed Count", title = "Distribution of Seed Count by Species and Treatment") +
  theme_minimal()
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 12. Distribution of Seed Count by Species and Treatment.** Boxplot showing the distribution of seed count by species and treatment. The x-axis represents different species, the y-axis represents seed count, and the fill color represents treatment (shrub or open).
:::

```{r results-vis-2}
# show the relationship of total.nr.infl and seeds
ggplot(data = seedmod3, aes(x = total_nr_infl, y = nr_seeds)) +
  geom_point() +
  labs(x = "Total Number of Inflorescences", y = "Seed Count", title = "Relationship between Seed Count and Total Number of Inflorescences") +
  theme_minimal()
```

::: {style="font-size: 90%; color: #787878;"}
**Figure 13. Relationship between Seed Count and Total Number of Inflorescence.** Scatter plot depicting the relationship between seed count and the total number of inflorescence. The x-axis represents the total number of inflorescence, and the y-axis represents seed count.
:::

```{r}
table <- tidy(seedmod3, conf.int = TRUE) %>% 
  # change the p-value numbers if they're really small using mutate
  # change the estimates, st err, and t-stats to round to ___ digits
  mutate(p.value = round(p.value, digits = 4),
         estimate = round(estimate, digits = 4),
         std.error = round(std.error, digits = 4)) %>% 
  # make it into a flextable
  flextable() %>% 
  # fit it to the viewer
  autofit()

table
```

::: {style="font-size: 90%; color: #787878;"}
**Table 1. Statistical results for variables as predictor of seed count.** The table summarizes the statistical results for the predictors in the model. "treatmentshrub" has a significant negative effect (p = 0.016), while "speciesCARRUP" and "total_nr_infl" have significant positive effects (p \< 0.001). "speciesTRIDAS" also has a significant positive effect (p = 0.042). Other predictors do not show significant effects. The table provides key information on the relationships between predictors and seed count
:::

### Interpretation of Results

These results indicated that seed count is significantly affected by plot type, species, and number of inflorescence. When these three variables are at play together, we are most able to predict seed type. Open plots were associated with more seed counts than shrub plots, as well as species Carrup and Tridas had much higher seed counts than other species. Total number of inflorescence also had a positive relationship with seed count, increasing mutually. Our results allow us to reject our null hypothesis that none of the variables would affect seed count, instead all had an impact. In an effort to see if any simpler models existed that could predict seed count, like only plot type, species, or number of inflorescence, but none of the models were better than the model that accounted for all the variables as the predictors. This tells us that all these variables should be accounted for in strategic planning for plant reproductive success, and that the ideal would be using an open plot, with either the Carrup or Tridas species, and having a higher number of inflorescence to have the highest seed count. The findings that shrub plots were less successful for seed counts may be cause for remedies to the 'shrubification' ocurring in the alpine tundra in order for overall plant community success to persevere.

## Works Cited

1.  Filazzola, A., Liczner, A. R., Westphal, M., & Lortie, C. J. (2019). Shrubs indirectly increase desert seedbanks through facilitation of the plant community. PloS one, 14(4), e0215988. <https://doi.org/10.1371/journal.pone.0215988>
2.  Mod, H. K., & Luoto, M. (2016). Arctic shrubification mediates the impacts of warming climate on changes to tundra vegetation. Environmental Research Letters, 11(12). DOI 10.1088/1748-9326/11/12/124028
3.  Paul-Victor, C., & Turnbull, L. A. (2009). The effect of growth conditions on the seed size/number trade-off. PloS one, 4(9), e6917. <https://doi.org/10.1371/journal.pone.0006917>
