---
title: "ENVS-193DS_kangaroo-rats"
author: Lena Hicks, Jasmine Toni, Izzy Tector
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: yeti
editor: visual
execute: 
  message: false
  warning: false
editor: visual
---

# Problem 1. Kangaroo Rats

## Introduction

Disturbance plays a vital biological role in shaping plant community structure, and small-scale disturbances caused by animals can increase heterogeneity and species diversity. Kangaroo rat (Dipodomys spectabilis) mounds are unique small-scale disturbances that support distinct plant communities compared to surrounding habitats (Guo). The seed bank, often overlooked in plant communities, is an important trait for plant species' persistence in unpredictable environments (Fitch). To understand the impact of kangaroo rat disturbances on plant communities in the Chihuahuan Desert grassland, we studied the composition of the seed bank for eight taxa on and off kangaroo rat mounds. Our objective was to determine if kangaroo rat mounds accumulate more seeds and exhibit different seed compositions compared to the adjacent grassland and our findings revealed that kangaroo rat mounds accumulated more seeds and supported different seed compositions, suggesting their role in promoting plant diversity and altering plant structure within communities (Koontz). Our null hypothesis is that the eight taxa will have seed banks that do not differ between mounds and off mounds and that mounds will not have high seed numbers of early successional plant species; our alternative hypothesis is that the eight taxa will have seed banks that differ between mounds and off mounds and that mounds will have high seed numbers of early successional plant species.

## Methods

```{r}
# should haves
library(MASS) 
library(lme4)
library(glmmTMB)
library(tidyverse)
library(DHARMa)
library(lmtest)
library(here)
library(janitor)
library(dplyr)
library(ggeffects)
library(performance)
library(naniar)
library(flextable)
library(car)
library(broom)
library(corrplot)
library(AICcmodavg)
library(GGally)
library(skimr)
# would be nice to have
library(MuMIn)
#library(equatiomatic)
library(corrplot)
```

```{r load-in-data}
kangaroo_rats <- read_csv(here("data", "dat_for_prob1.csv")) %>% 
  # make the column names cleaner
  clean_names()
```

```{r}
gg_miss_var(kangaroo_rats)
```

```{r}
skim(kangaroo_rats)
```

```{r}
# Cleaning and wrangling
data <- kangaroo_rats %>%
  drop_na()    # Remove rows with missing values

# Assumption checks
# Checking normality assumption
shapiro.test(data$seeds)

# Checking homogeneity of variances assumption
leveneTest(seeds ~ loc, data = data)

# Statistical tests
# One-way ANOVA
anova_result <- aov(seeds ~ loc, data = data)
summary(anova_result)

# Post hoc test (Tukey's HSD)
tukey_result <- TukeyHSD(anova_result)
tukey_result

# Null hypotheses
# H0: There is no difference in seed counts between locations

# Other tools for count (discrete) data:
# Kruskal-Wallis test (non-parametric alternative to ANOVA)
kruskal.test(seeds ~ loc, data = data)
```

## Results

these are the results portion

## Works Cited

Fitch, Henry S. \"Habits and Economic Relationships of the Tulare Kangaroo Rat.\" *Journal of Mammalogy* 29, no. 1 (1948): 5--35. https://doi.org/10.2307/1375277.

Guo, Q. Effects of bannertail kangaroo rat mounds on small-scale plant community structure. *Oecologia* **106**, 247--256 (1996). https://doi.org/10.1007/BF00328605

Koontz, Terri, and Heather Simpson. \"The Composition of Seed Banks on Kangaroo Rat (Dipodomys Spectabilis) Mounds in a Chihuahuan Desert Grassland.\" ScienceDirect, 18 Apr. 2010, doi.org/10.1016/j.jaridenv.2010.03.008.

\
